Accelerating Optimal Power Flow with GPUs: SIMD Abstraction of Nonlinear Programs and Condensed-Space Interior-Point Methods

### 2023-07-28 paper/main.tex
before merging with Mihai's edit

While derivative evaluation can be generally cheaper than linear
algebra operations, our numerical results on AC OPF problems show that
AD often constitutes more than half of the total solver time when
using off-the-shelf AD implementations like
JuMP.jl~\cite{dunning2017jump} or
AMPL~\cite{fourer1990modeling}. Instead, our method leverages a
specialized AD implementation based on the SIMD abstraction of
NLPs. This abstraction allows to preserve the parallelizable structure
within the model equations, facilitating efficient derivative
evaluations on the GPU. The AC power flow model is particularly
well-suited for this abstraction as it involves repetitive expressions
for each component type (e.g., buses, lines, generators), and the
number of computational patterns does not increase with the network's
size. Numerical results reported in this paper demonstrate that our
proposed strategies can achieve over 20 times speedup by running on
GPU. In comparison to general AD implementations on CPUs (such as AMPL
and JuMP.jl), our GPU-based  differentiation method can be approximately 50 times faster.

@@

\paragraph*{Contributions}
We present, for the first time, a nonlinear optimization framework
that can run entirely on GPU, with all the performance-critical data
arrays residing exclusively on GPUs. Additionally, we introduce the
concept of SIMD abstraction for NLP problems, which results in an
efficient implementation of GPU-accelerated parallel AD, leading to a
50x speedup compared to existing tools for large-scale
OPFs. Furthermore, we propose the condensed IPM with inequality
relaxation strategy for the first time, enabling the treatment of the
KKT systems of NLPs without numerical pivoting, thus allowing the
solution of sparse, large-scale NLPs (with a prominent example being
AC OPFs) on GPUs.

@@

At the beginning of the algorithm, we apply inequality relaxation to replace the equality constraints in \eqref{eqn:cpt} with inequalities by introducing slack variables $s\in\mathbb{R}^{m}$:
\begin{align}\label{eqn:relax}
  g(x)- s = 0,\quad s^{\flat}\leq s\leq  s^\sharp,
\end{align}
where $s^\flat,s^\sharp\in\mathbb{R}^{m}$ are lower and upper bounds chosen to be close to zero.
This relaxed problem can be stated as follows:
\begin{equation}\label{eqn:relaxed}
    \min_{\big[\substack{x^\flat\\s^\flat}\big]\leq \big[\substack{x\\s}\big] \leq \big[\substack{x^\sharp\\s^\sharp}\big]}\;
      f(x)\quad \st\;  g(x) - s = 0 \, .
\end{equation}
In our implementation, we set $s^{\flat},s^\sharp$ as
$-\epsilon_{\text{tol}}\boldsymbol{1}$ and $+\epsilon_{\text{tol}}\boldsymbol{1}$, respectively,
where $\epsilon_{\text{tol}}>0$ is a user-specified relative tolerance
of the IPM. This type of relaxation is
commonly used in practical IPM implementations; for example, in Ipopt,
the solver relaxes the bounds and inequality constraints by
$O(\epsilon_{\text{tol}})$ to prevent an empty interior of the
feasible set (see \cite[Section 3.5]{wachter2006implementation}).
For condensed-space IPM, we cannot maintain the same level of precision
due to the increased condition number of the KKT system.
We have found that setting $\epsilon_{\text{tol}}$ to be
$\epsilon_{\text{mach}}^{1/4}\approx 10^{-4}$ ensures numerical
stability while achieving satisfactory convergence behavior. Thus, our
solver sets the tolerance to $10^{-4}$ by default when using condensed
IPM.

@@

\subsection{Newton's Step Computation}
The Newton step direction is computed by solving a so-called KKT system; to explain this, we consider the first-order optimality conditions (KKT conditions) for the barrier subproblem in \eqref{eqn:barrier}:
\begin{align}\label{eqn:first}
  \begin{aligned}[t]
    \nabla_{x} f(x) - \nabla_{x}g(x)^\top y  - z_x^\flat  + z_x^\sharp = 0\;&\\
    \begin{aligned}
      - z_s^\flat  + z_s^\sharp  &= 0\\
      Z^\flat_x (x-x^\flat) - \mu\boldsymbol{1} &= 0\\
      Z^\flat_s (s-s^\flat) - \mu\boldsymbol{1}&= 0,
    \end{aligned}
    \quad
    \begin{aligned}
      g(x) - s  &= 0\\
      Z^\sharp_x (x^\sharp-x) - \mu\boldsymbol{1}&= 0\\
      Z^\sharp_s (s^\sharp-s) - \mu\boldsymbol{1}&= 0,
    \end{aligned}&
  \end{aligned}
\end{align}
where $y\in\mathbb{R}^{m}$, $z_x^\flat,z_x^\sharp\in\mathbb{R}^{n}$,
and $z_s^\flat,z_s^\sharp\in\mathbb{R}^{m}$ are Lagrange multipliers
associated with the equality and bound constraints in
\eqref{eqn:relaxed}. The Newton step for solving the nonlinear
equations in \eqref{eqn:first} can be computed by solving the unreduced KKT system in \eqref{eqn:very-long-eqn}.
\begin{figure*}[h!]
  % \begin{strip}
  % \hline
  \begin{align}\label{eqn:very-long-eqn}
    \underbrace{
    \begin{bmatrix}
      W^{(\ell)}  + \delta^{(\ell)}_w I & & A^{(\ell)\top}& -I & I &  \\
      & \delta^{(\ell)}_w I & -I&&&-I & I\\
      A^{(\ell)}& -I & -\delta^{(\ell)}_c I\\
      Z_x^{(\ell)\flat}&&&X^{(\ell)}-X^\flat\\
      -Z_x^{(\ell)\sharp}&&&&X^\sharp-X^{(\ell)}\\
      &Z_s^{(\ell)\flat}&&&&S^{(\ell)}-S^\flat\\
      &-Z_s^{(\ell)\sharp}&&&&&S^\sharp-S^{(\ell)}\\
    \end{bmatrix}
    }_{M_\text{full}}
    \begin{bmatrix}
      \Delta x \\
      \Delta s \\
      \Delta y \\
      \Delta z_x^\flat \\
      \Delta z_x^\sharp \\
      \Delta z_s^\flat \\
      \Delta z_s^\sharp \\
    \end{bmatrix} =
    \begin{bmatrix}
      p^{(\ell)}_{x }\\
      p^{(\ell)}_{s }\\
      p^{(\ell)}_{y }\\
      p^{(\ell)}_{z_x^\flat }\\
      p^{(\ell)}_{z_x^\sharp }\\
      p^{(\ell)}_{z_s^\flat }\\
      p^{(\ell)}_{z_s^\sharp }\\
    \end{bmatrix}
  \end{align}
% \hline
% \end{strip}
\end{figure*}
Here, we recall the definitions of $W^{(\ell)}$ and $A^{(\ell)}$ from Section \ref{sec:numopt}, and
$p^{(\ell)}_x,\cdots p^{(\ell)}_{z_s^\sharp}$ are defined by the left-hand-sides of the equations in
\eqref{eqn:first}. In sequel, we shall drop the superscript
$(\cdot)^{(\ell)}$ for concise notation.

Now, we observe that a significant portion of the system in
\eqref{eqn:very-long-eqn} can be eliminated by exploiting the block
structure, leading to an equivalent system stated in a smaller
space. In particular, the lower-right $4\times 4$ block is always
invertible since the IPM procedure ensures that the iterates stay in
the strict interior of the feasible set. This allows for eliminating
the lower-right 4x4 block, resulting in:
\begin{align}\label{eqn:very-long-reduced}
  &
    \underbrace{
    \begin{bmatrix}
      W  + \Sigma_x + \delta_w I & & A^{\top} \\
      & \Sigma_s + \delta_w I & -I\\
      A& -I & -\delta_c I\\
    \end{bmatrix}
  }_{M_{\text{red}}}
  \begin{bmatrix}
    \Delta x \\
    \Delta s \\
    \Delta y \\
  \end{bmatrix}=
  \begin{bmatrix}
    q_x \\
    q_s\\
    q_y\\
  \end{bmatrix},
\end{align}
where the dependence on the evaluation point is suppressed for concise notation, and:
\begin{align*}
  \Sigma_x&:= Z^\flat_x (X-X^\flat)^{-1}+ Z^\sharp_x (X^\sharp-X)^{-1}\\
  \Sigma_s&:= Z^\flat_s (S-S^\flat)^{-1}+ Z^\sharp_s (S^\sharp-S)^{-1}\\
  q_x&:=p_x + (X-X^\flat)^{-1} p_{z^\flat_x}-  (X^\sharp-X)^{-1} p_{z^\sharp_x}\\
  q_s&:= (S-S^\flat)^{-1} p_{z^\flat_s}-  (S^\sharp-S)^{-1} p_{z^\sharp_s}\\
  q_y&:=p_y.
\end{align*}
The bound dual steps can be recovered as follows:
\begin{align}\label{eqn:recover-1}
  \begin{aligned}[t]
    \Delta z^\flat_x &= \left(X-X^\flat\right)^{-1} \left(-Z^\flat_x \Delta x  + p_{z^\flat_x}\right)\\
    \Delta z^\sharp_x &= \left(X^\sharp-X\right)^{-1} \left(Z^\sharp_x \Delta x  + p_{z^\sharp_x}\right)\\
    \Delta z^\flat_s &= \left(S-S^\flat\right)^{-1} \left(-Z^\flat_s \Delta s  + p_{z^\flat_s}\right)\\
    \Delta z^\sharp_s &= \left(S^\sharp-S\right)^{-1} \left(Z^\sharp_s \Delta s  + p_{z^\sharp_s}\right).
  \end{aligned}
\end{align}
Note that the matrices involved in the inversions in \eqref{eqn:recover-1}
are always diagonal, so their computation is cheap.
Also, note that the reduced system in \eqref{eqn:very-long-reduced}
corresponds to the KKT system in \eqref{eqn:kkt-indefinite}. However,
in the original version of the algorithm, we did not introduce the
slack variables, so it does not have the additional structure imposed
by the slack variables.

The key advantage of the inequality relaxation strategy is that it
imposes additional structure to the reduced KKT system, allowing us to
further reduce the dimension of the problem. In particular, the
lower-right 2x2 block in \eqref{eqn:very-long-reduced} can be
eliminated, which is a procedure called {\it condensation}; here, the invertibility of the lower-right block can be verified from the fact that $\delta_w,\delta_c\geq 0$ and $\Sigma_s\succ 0$. Through this,
we obtain the following system, written in the  primal
space only:
\begin{align}\label{eqn:very-long-condensed}
  (\underbrace{W + \delta_wI + \Sigma_x + A^{\top} D A}_{M_\text{cond}} ) \Delta x = q_x + A^\top (C q_y +  Dq_s ),
\end{align}
where:
\begin{align*}
  C := \left(\delta_c \Sigma_s + (1+\delta^{}_c\delta^{}_w) I\right)^{-1}, \;
  D := \left(\Sigma_s + \delta^{}_w I\right)C,
\end{align*}
and the dual and slack step directions can be recovered by:
\begin{align}
  \Delta s &:= C \left(\delta_c q_s - (q_y + A\Delta x)\right)\nonumber\\
  \Delta y &:= (\Sigma_s + \delta_w I) \Delta s -q_s.\label{eqn:recover-2}
\end{align}
Again, the matrices involved in the inversions above are always
diagonal, so their computation is cheap.

Therefore, the only sparse matrix that needs to be factorized is the
matrix in the left-hand-side of \eqref{eqn:very-long-condensed}, with
dimension $n \times n$.
In general NLPs, the condensation strategy
can arbitrarily increase the density of the KKT system, but for AC OPF
problems, the condensed KKT system remains sparse.
This is because the maximum number of nonzeros per row in constraint
Jacobian is bounded by the graph degree, which should be reasonably
small in practice.

@@

Although we call \eqref{eqn:very-long-condensed} a condensed KKT system,
$M_\text{cond}$ is not necessarily a dense matrix. In fact, in the case of
AC OPF problems, $M_\text{cond}$ is still highly sparse, as $W$
and $A$ are graph-induced banded systems. Thus, exploiting sparsity is
still necessary to enable scalable computations.

@@

The numerical benchmark results, including total solution time and its
breakdown into linear algebra and derivative evaluation time (with the
remainder considered as solver internal time), are shown in Table
\ref{tbl:results}. The quality of the solution (objective value and
constraint violation measured by $\|\cdot\|_\infty$) is shown in Table
\ref{tbl:quality}. Figure \ref{fig:speedup} visually represents the
speedup brought by GPUs, by comparing the timing results of
\ref{config-1} and \ref{config-2}.

@@

\paragraph{Convergence pattern}
First, when comparing the solvers' performance in terms of the
interior point method (IPM) iteration counts, MadNLP.jl is as
efficient as the state-of-the-art solver Ipopt. The IPM iteration
count is nearly the same as that of Ipopt for achieving the same level
of accuracy in the final solution (see Table \ref{tbl:quality}). This
suggests that running mathematically equivalent operations on GPUs (by
using shared code base in high-level abstractions) can yield a similar
degree of effectiveness in terms of IPM convergence.

\paragraph{Performance of SIMD AD}
Next, we discuss the effectiveness of parallel AD on GPUs. We observe
that even on CPUs, SIMDiff.jl is substantially faster than the AD
routines implemented in AMPL or JuMP.jl. Indeed,
SIMDiff.jl generates derivative functions specifically compiled
for the type of model, including optimization for the distinctive
computational pattern found in the model. When comparing SIMDiff.jl running
on CPUs and GPUs, we observe a further speedup on the GPU of up to almost 20x for
large instances (e.g., case24465\_goc); remarkably, this is 50 times
faster than JuMP.jl. This demonstrates that parallelizing AD
brings significant computational gain.
This is because most operations in the derivative
evaluations are \ref{eqn:pattern-1}
operations, highly effective on GPUs.

\paragraph{Performance of the condensed KKT strategy}
We next discuss the linear solver time. While the speedup achieved by
linear solvers is only moderate, this has a high impact on the overall
speedup, as linear solver time constitutes a significant portion of
the total solution time. For large instances, approximately 4x speedup
can be achieved, though this can vary depending on the instances. For
example, for case24464\_goc, the GPU linear solver was slower than the
CPUs. The investigation of under which circumstances {\tt cuSOLVERRF} is
more effective warrants further research.

Solver internal time could also be significantly accelerated through
GPU utilization. We can observe that the speedup in solver internal
operations is consistently greater than the speedup in linear
solvers. However, due to the frequent use of \ref{eqn:pattern-2}
operations, the speedup in solver internal operations is lesser than
that of derivative evaluations.

Overall, our GPU implementation exhibits significant speedup across
all components (derivative evaluation, linear algebra, and solver
internal computation) resulting in substantial gains in total solution
time. The results indicate that GPUs become more effective for
large-scale instances, particularly when the number of variables is
greater than 20,000. Notably, for the largest instance,
case30000\_goc, our GPU implementation is 4 times faster than our CPU
implementation and approximately 10 times faster than
state-of-the-art tools (Ipopt, JuMP.jl, and Ma27).
This demonstrates that GPUs can bring significant computational
gains for large-scale AC OPF problems, enabling the solution of
previously inconceivable problems due to the limitations of existing
CPU-based solution tools.

>>>>

While derivative evaluation can be generally cheaper than linear
algebra operations, our numerical results on AC OPF problems show that
AD often constitutes more than half of the total solver time when
using off-the-shelf AD implementations like
JuMP.jl~\cite{dunning2017jump} or
AMPL~\cite{fourer1990modeling}. Instead, our method leverages a
specialized AD implementation based on the SIMD abstraction of
NLPs. This abstraction allows to preserve the parallelizable structure
within the model equations, facilitating efficient derivative
evaluations on the GPU. The AC power flow model is particularly
well-suited for this abstraction as it involves repetitive expressions
for each component type (e.g., buses, lines, generators), and the
number of computational patterns does not increase with the network's
size. Numerical results reported in this paper demonstrate that our
proposed strategies can achieve over 20 times speedup by running on
GPU. In comparison to general AD implementations on CPUs (such as AMPL
and JuMP.jl), our GPU-based differentiation method can be
approximately 500 times faster.

@@

\paragraph*{Contributions}
We present, for the first time, a nonlinear optimization framework
that can run entirely on GPU, with all the performance-critical data
arrays residing exclusively on GPUs. Additionally, we introduce the
concept of SIMD abstraction for NLP problems, which results in an
efficient implementation of GPU-accelerated parallel AD. Furthermore,
we propose the condensed IPM with inequality relaxation strategy for
the first time, enabling the treatment of the KKT systems of NLPs
without numerical pivoting, thus allowing the solution of sparse,
large-scale NLPs (with a prominentexample being AC OPFs) on GPUs.

@@

At the beginning of the algorithm, we apply inequality relaxation to replace the equality constraints in \eqref{eqn:cpt} with inequalities by introducing slack variables $s\in\mathbb{R}^{m}$:
\begin{align}\label{eqn:relax}
  g(x)- s = 0,\quad s^{\flat}\leq s\leq  s^\sharp,
\end{align}
where $s^\flat,s^\sharp\in\mathbb{R}^{m}$ are lower and upper bounds chosen to be close to zero.
This relaxed problem can be stated as follows:
\begin{equation}\label{eqn:relaxed}
    \min_{\big[\substack{x^\flat\\s^\flat}\big]\leq \big[\substack{x\\s}\big] \leq \big[\substack{x^\sharp\\s^\sharp}\big]}\;
      f(x)\quad \st\;  g(x) - s = 0 \, .
\end{equation}
In our implementation, we set $s^{\flat},s^\sharp$ as
$-\epsilon_{\text{tol}}\boldsymbol{1}$ and $+\epsilon_{\text{tol}}\boldsymbol{1}$, respectively,
where $\epsilon_{\text{tol}}>0$ is a user-specified relative tolerance
of the IPM. This type of relaxation is
commonly used in practical IPM implementations; for example, in Ipopt,
the solver relaxes the bounds and inequality constraints by
$O(\epsilon_{\text{tol}})$ to prevent an empty interior of the
feasible set (see \cite[Section 3.5]{wachter2006implementation}).
For condensed-space IPM, we cannot maintain the same level of precision
due to the increased condition number of the KKT system.
We have found that setting $\epsilon_{\text{tol}}$ to be
$\epsilon_{\text{mach}}^{1/4}\approx 10^{-4}$ ensures numerical
stability while achieving satisfactory convergence behavior. Thus, our
solver sets the $\epsilon_\text{tol}$ to $10^{-4}$ by default when using condensed
IPM.

@@

\subsection{Newton's Step Computation}
The Newton step direction is computed by solving a so-called KKT system; to explain this, we consider the first-order optimality conditions (KKT conditions) for the barrier subproblem in \eqref{eqn:barrier}:
\begin{align}\label{eqn:first}
  \begin{aligned}[t]
    \nabla_{x} f(x) - \nabla_{x}g(x)^\top y  - z_x^\flat  + z_x^\sharp = 0\;&\\
    \begin{aligned}
      - z_s^\flat  + z_s^\sharp + y &= 0,\\
      Z^\flat_x (x-x^\flat) - \mu\boldsymbol{1} &= 0,\\
      Z^\flat_s (s-s^\flat) - \mu\boldsymbol{1}&= 0,
    \end{aligned}
    \quad
    \begin{aligned}
      g(x) - s  &= 0\\
      Z^\sharp_x (x^\sharp-x) - \mu\boldsymbol{1}&= 0\\
      Z^\sharp_s (s^\sharp-s) - \mu\boldsymbol{1}&= 0,
    \end{aligned}&
  \end{aligned}
\end{align}
where $y\in\mathbb{R}^{m}$, $z_x^\flat,z_x^\sharp\in\mathbb{R}^{n}$,
and $z_s^\flat,z_s^\sharp\in\mathbb{R}^{m}$ are Lagrange multipliers
associated with the equality and bound constraints in
\eqref{eqn:relaxed}. The Newton step for solving the nonlinear
equations in \eqref{eqn:first} can be computed by solving the system in Fig. \ref{fig:very-long-eqn}.
\begin{figure*}[t]
  % \begin{strip}
  % \hline
  \begin{align*}
    \underbrace{
    \begin{bmatrix}
      W^{(\ell)}  + \delta^{(\ell)}_w I & & A^{(\ell)\top}& -I & I &  \\
      & \delta^{(\ell)}_w I & -I&&&-I & I\\
      A^{(\ell)}& -I & -\delta^{(\ell)}_c I\\
      Z_x^{(\ell)\flat}&&&X^{(\ell)}-X^\flat\\
      -Z_x^{(\ell)\sharp}&&&&X^\sharp-X^{(\ell)}\\
      &Z_s^{(\ell)\flat}&&&&S^{(\ell)}-S^\flat\\
      &-Z_s^{(\ell)\sharp}&&&&&S^\sharp-S^{(\ell)}\\
    \end{bmatrix}
    }_{M_\text{full}}
    \begin{bmatrix}
      \Delta x \\
      \Delta s \\
      \Delta y \\
      \Delta z_x^\flat \\
      \Delta z_x^\sharp \\
      \Delta z_s^\flat \\
      \Delta z_s^\sharp \\
    \end{bmatrix} =
    \begin{bmatrix}
      p^{(\ell)}_{x }\\
      p^{(\ell)}_{s }\\
      p^{(\ell)}_{y }\\
      p^{(\ell)}_{z_x^\flat }\\
      p^{(\ell)}_{z_x^\sharp }\\
      p^{(\ell)}_{z_s^\flat }\\
      p^{(\ell)}_{z_s^\sharp }\\
    \end{bmatrix}
  \end{align*}
  \caption{Full KKT system}\label{fig:very-long-eqn} 
\end{figure*}
Here, we recall the definitions of $W^{(\ell)}$ and $A^{(\ell)}$ from Section \ref{sec:numopt}, and
$p^{(\ell)}_x,\cdots p^{(\ell)}_{z_s^\sharp}$ are defined by the left-hand-sides of the equations in
\eqref{eqn:first}. In sequel, we shall drop the superscript
$(\cdot)^{(\ell)}$ for concise notation.

Now, we observe that a significant portion of the system in Fig.
\eqref{fig:very-long-eqn} can be eliminated by exploiting the block
structure, leading to an equivalent system stated in a smaller
space. In particular, the lower-right $4\times 4$ block is always
invertible since the IPM procedure ensures that the iterates stay in
the strict interior of the feasible set. This allows for eliminating
the lower-right 4x4 block, resulting in:
\begin{align}\label{eqn:very-long-reduced}
  &
    \underbrace{
    \begin{bmatrix}
      W  + \Sigma_x + \delta_w I & & A^{\top} \\
      & \Sigma_s + \delta_w I & -I\\
      A& -I & -\delta_c I\\
    \end{bmatrix}
  }_{M_{\text{red}}}
  \begin{bmatrix}
    \Delta x \\
    \Delta s \\
    \Delta y \\
  \end{bmatrix}=
  \begin{bmatrix}
    q_x \\
    q_s\\
    q_y\\
  \end{bmatrix},
\end{align}
where:
\begin{align*}
  \Sigma_x&:= Z^\flat_x (X-X^\flat)^{-1}+ Z^\sharp_x (X^\sharp-X)^{-1}\\
  \Sigma_s&:= Z^\flat_s (S-S^\flat)^{-1}+ Z^\sharp_s (S^\sharp-S)^{-1}\\
  q_x&:=p_x + (X-X^\flat)^{-1} p_{z^\flat_x}-  (X^\sharp-X)^{-1} p_{z^\sharp_x}\\
  q_s&:= (S-S^\flat)^{-1} p_{z^\flat_s}-  (S^\sharp-S)^{-1} p_{z^\sharp_s}\\
  q_y&:=p_y.
\end{align*}
The bound dual steps can be recovered as follows:
\begin{align}\label{eqn:recover-1}
  \begin{aligned}[t]
    \Delta z^\flat_x &= \left(X-X^\flat\right)^{-1} \left(-Z^\flat_x \Delta x  + p_{z^\flat_x}\right)\\
    \Delta z^\sharp_x &= \left(X^\sharp-X\right)^{-1} \left(Z^\sharp_x \Delta x  + p_{z^\sharp_x}\right)\\
    \Delta z^\flat_s &= \left(S-S^\flat\right)^{-1} \left(-Z^\flat_s \Delta s  + p_{z^\flat_s}\right)\\
    \Delta z^\sharp_s &= \left(S^\sharp-S\right)^{-1} \left(Z^\sharp_s \Delta s  + p_{z^\sharp_s}\right).
  \end{aligned}
\end{align}
Note that the matrices involved in the inversions in \eqref{eqn:recover-1}
are always diagonal, so their computation is cheap.
Also, note that the reduced system in \eqref{eqn:very-long-reduced}
corresponds to the KKT system in \eqref{eqn:kkt-indefinite}. However,
in the original version of the algorithm, we did not introduce the
slack variables, so it does not have the additional structure imposed
by the slack variables.

The key advantage of the inequality relaxation strategy is that it
imposes additional structure to the augmented KKT system, allowing us to
further reduce the dimension of the problem. In particular, the
lower-right 2x2 block in \eqref{eqn:very-long-reduced} can be
eliminated, which is a procedure called {\it condensation}; here, the invertibility of the lower-right block can be verified from the fact that $\delta_w,\delta_c\geq 0$ and $\Sigma_s\succ 0$. Through this,
we obtain the following system, written in the  primal
space only:
\begin{align}\label{eqn:very-long-condensed}
  (\underbrace{W + \delta_wI + \Sigma_x + A^{\top} D A}_{M_\text{cond}} ) \Delta x = q_x + A^\top (C q_y +  Dq_s ),
\end{align}
where:
\begin{align*}
  C := \left(\delta_c \Sigma_s + (1+\delta^{}_c\delta^{}_w) I\right)^{-1}, \;
  D := \left(\Sigma_s + \delta^{}_w I\right)C,
\end{align*}
and the dual and slack step directions can be recovered by:
\begin{align}
  \Delta s &:= C \left(\delta_c q_s - (q_y + A\Delta x)\right)\nonumber\\
  \Delta y &:= (\Sigma_s + \delta_w I) \Delta s -q_s.\label{eqn:recover-2}
\end{align}
Again, the matrices involved in the inversions above are always
diagonal, so their computation is cheap.

Therefore, the only sparse matrix that needs to be factorized is the
matrix in the left-hand-side of \eqref{eqn:very-long-condensed}, with
dimension $n \times n$.  Although we call
\eqref{eqn:very-long-condensed} a condensed KKT system,
$M_\text{cond}$ is not necessarily a dense matrix. In fact, in the
case of AC OPF problems, $M_\text{cond}$ is still highly sparse, as
$W$ and $A$ are graph-induced banded systems.  Thus, exploiting
sparsity is still necessary to enable scalable computations.  In
general NLPs, however, the condensation strategy can arbitrarily
increase the density of the KKT system. Thus, the condensed-space IPM
strategy needs to be used with caution.

@@

@@

The numerical benchmark results, including total solution time and its
breakdown into linear algebra and derivative evaluation time (with the
remainder considered as solver internal time), are shown in Table
\ref{tbl:results}. The quality of the solution (objective value and
constraint violation measured by $\|\cdot\|_\infty$) is shown in Table
\ref{tbl:quality}. Fig. \ref{fig:speedup} visually represents the
speedup brought by GPUs, by comparing the timing results of
\ref{config-1} and \ref{config-2}.

@@

\paragraph*{Convergence pattern}
First, when comparing the solvers' performance in terms of the
interior point method (IPM) iteration counts, MadNLP.jl is as
efficient as the state-of-the-art solver Ipopt. The IPM iteration
count is nearly the same as that of Ipopt for achieving the same level
of accuracy in the final solution (see Table \ref{tbl:quality}). This
suggests that running mathematically equivalent operations on GPUs (by
using shared code base in high-level abstractions) can yield a similar
degree of effectiveness in terms of IPM convergence.

\paragraph*{Performance of AD}
Next, we discuss the effectiveness of parallel AD on GPUs. We observe
that even on CPUs, SIMDiff.jl is substantially faster than the AD
routines implemented in AMPL or JuMP.jl. Indeed, SIMDiff.jl generates
derivative functions specifically compiled for the type of model,
including optimization for the distinctive computational pattern found
in the model. When comparing SIMDiff.jl running on CPUs and GPUs, we
observe a further speedup on the GPU of up to almost 20x for large
instances (e.g., case24465\_goc); remarkably, this is 500 times faster
than JuMP.jl. This demonstrates that adopting SIMD abstraction and
parallelizing AD brings significant computational gain.

\paragraph*{Performance of the Condensed-Space IPM}
We next discuss the linear solver time. While the speedup achieved by
linear solvers is only moderate, this has a high impact on the overall
speedup, as linear solver time constitutes a significant portion of
the total solution time. For large instances, approximately 4x speedup
can be achieved, though this can vary depending on the instances. For
example, for case24464\_goc, the GPU linear solver was slower than the
CPUs. The investigation of under which circumstances {\tt cuSOLVERRF} is
more effective warrants further research.

Solver internal time could also be significantly accelerated through
GPU utilization. We can observe that the speedup in solver internal
operations is consistently greater than the speedup in linear
solvers. However, due to the frequent use of \ref{eqn:pattern-2} and
\ref{eqn:pattern-3} operations, the speedup in solver internal
operations is lesser than that of derivative evaluations.

Overall, our GPU implementation exhibits significant speedup across
all components (derivative evaluation, linear algebra, and solver
internal computation) resulting in substantial gains in total solution
time. The results indicate that GPUs become more effective for
large-scale instances, particularly when the number of variables is
greater than 20,000. Notably, for the largest instance,
case30000\_goc, our GPU implementation is 4 times faster than our CPU
implementation and approximately 10 times faster than
state-of-the-art tools (Ipopt, JuMP.jl, and Ma27).
This demonstrates that GPUs can bring significant computational
gains for large-scale AC OPF problems, enabling the solution of
previously inconceivable problems due to the limitations of CPU-based
solution tools.

\paragraph*{Known Limiataions}
The key limitation of our method is the decreased precision of the
final solution. Reliable convergence is achieved only up to a
tolerance of $10^{-4}$. We have observed that the condensation of the
KKT system worsens the conditioning of the already ill-conditioned
augmented KKT system, resulting in even higher condition numbers,
particularly when the solution is almost converged. For instance, in
the case of case118\_ieee, the condition number of the KKT system at
the solution (with tolerance set to $10^{-4}$) is $9.43\times 10^{11}$
for the augmented KKT system and $3.15\times 10^{14}$ for the
condensed KKT system. As a consequence, the achievable precision of
the solution is reduced. Further investigation into the possibility of
obtaining higher precision will be necessary in the future.
